Does stomatal patterning in amphistomatous leaves minimize the CO\(_2\) diffusion path length with leaves of \emph{Arabidopsis thaliana}?

### 2023-10-19 ms/ms.Rmd
fixed scale following Graham's email on 2023-10-19

---
title: Does stomatal patterning in amphistomatous leaves minimize the CO$_2$ diffusion path length with leaves of *Arabidopsis thaliana*?
date: "`r Sys.Date()`"
author:
  - name: Jacob L. Watts
    email: Jacob.Watts-1@colorado.edu
    corresponding_author: yes # Footnote using corresponding_author
    affiliation: 
      - UHM
      - CU
  - name: Graham Dow
    email: graham.dow@niab.com
    affiliation: NIAB
  - name: Thomas N. Buckley
    email: tnbuckley@ucdavis.edu
    affiliation: UCD
  - name: Christopher D. Muir
    email: cdmuir@wisc.edu
    affiliation: 
      - UHM
      - UWM
    # footnote: 1 # Footnote using reference
address:
  - code: UHM
    address: School of Life Sciences, University of Hawaiʻi at Mānoa, Honolulu, HI 96822
  - code: CU
    address: Ecology and Evolutionary Biology, Universtity of Colorado, Boulder, CO 80309
  - code: NIAB
    address: National Institute of Agricultural Botany, Cambridge, United Kingdom CB3 0LE
  - code: UCD
    address: Department of Plant Sciences, University of California, Davis, CA 95616
  - code: UWM
    address: Department of Botany, University of Wisconsin, Madison, WI 53706
# footnote:
#   - code: 1
#     text: "Current email address: \\href{mailto:cdmuir@wisc.edu}{cdmuir@wisc.edu}"
abstract:
  - "We will write abstract after Discussion and Conclusions are complete"
keywords:
  - amphistomy
  - Arabidopsis thaliana
  - CO$_2$ diffusion
  - finite element method
  - optimality
  - photosynthesis
  - stomata
fontsize: 12pt
competing_interests: |
  The authors declare no competing interests.
author_contributions: |
  JLW and CDM conceived of the project, analyzed data, and wrote the manuscript. GW provided data. TNB contributed to model development and helped edit the manuscript.
acknowledgements: |
  This work was supported by NSF Award 1929167 to CDM and NSF Award 2307341 to TNB.
bibliography: stomata-spacing.bib
link-citations: yes
output: 
  rticles::oup_article:
    latex_engine: xelatex
    oup_version: 1 # 1 = 2020 CTAN OUP CLS package; 0 = 2009 OUP CLS package
    # journal: "AoB PLANTS"
    document_style: "modern" # Can be contemporary, modern, traditional
    papersize: "large" # Can be large, medium, small
    #citation_package: "default" # Uncomment when using a CSL; default "natbib"
    namedate: TRUE # Set FALSE to use numeric refs; Default FALSE
    number_sections: FALSE # Uncomment to not number sections; default TRUE
    #number_lines: TRUE # Use `lineno` package - Default FALSE
    #number_lines_options: ["mathlines","switch"]  # Options for latex lineno package.
    onecolumn: TRUE # Uncomment for one column format; default FALSE
    extra_dependencies:
      - booktabs # to use with knitr::kable() example below
header-includes:
 - \usepackage[nomarkers,tablesfirst]{endfloat} # For figures and tables at end
 - \usepackage{lineno} # For line numbering
 - \linenumbers # For line numbering
 - \usepackage{hyperref} # For autoref
 - \renewcommand{\figureautorefname}{Fig.}
 - \usepackage[detect-none]{siunitx} # for \numrange{}{}
 - \sisetup{range-phrase = \text{--}}
 - \usepackage{caption} # for \captionsetup{}
 - \usepackage{newunicodechar,graphicx}
 - \DeclareRobustCommand{\okina}{\raisebox{\dimexpr\fontcharht\font`A-\height}{\scalebox{0.8}{`}}}
 - \newunicodechar{ʻ}{\okina}

@@

@@

[Graham - can you fill in details on growth conditions, image aquisitiion, and image dimensions?]

*Arabidopsis thaliana* plants were grown in three different light environments: low light (PAR = 50 $\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$), medium light (100 $\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$), and high light (200 $\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$). PAR stands for photosynthetically active radiation. Once leaves were mature, we captured images of the abaxial and adaxial leaf surfaces using XXX. We captured 132 images in total, making 66 abaxial-adaxial image pairs. We measured stomatal position and size using ImageJ [@schneider_nih_2012].

@@

Stomatal density of *Arabidopsis thaliana* the `r nrow(single_surface_results)` leaf surfaces measured range from 12 to 93 (units) with high light leaves ranging from 93 to 55 (units), medium light from 15 to 35 (units), and low light from 12 to 42 (units). Stomatal density varies among light treatments (ANOVA, $`r report_fstat(density_anova, 'light')`$) because the density is much greater in the high light treatment (\autoref{fig:density}). Density is consistently greater on abaxial leaf surfaces across all light treatments (ANOVA, $`r report_fstat(density_anova, 'surface')`$; \autoref{fig:density}). There is no evidence for an interaction between light treatment and surface (ANOVA, $`r report_fstat(density_anova, 'light:surface')`$). Leaves are amphistomatous with a mean stomatal density ratio of 0.45.

>>>>

---
title: Does stomatal patterning in amphistomatous leaves minimize the CO$_2$ diffusion path length with leaves of *Arabidopsis thaliana*?
date: "`r Sys.Date()`"
author:
  - name: Jacob L. Watts
    email: Jacob.Watts-1@colorado.edu
    corresponding_author: yes # Footnote using corresponding_author
    affiliation: 
      - UHM
      - CU
  - name: Graham J. Dow
    email: graham.dow@niab.com
    affiliation: NIAB
  - name: Thomas N. Buckley
    email: tnbuckley@ucdavis.edu
    affiliation: UCD
  - name: Christopher D. Muir
    email: cdmuir@wisc.edu
    affiliation: 
      - UHM
      - UWM
    # footnote: 1 # Footnote using reference
address:
  - code: UHM
    address: School of Life Sciences, University of Hawaiʻi at Mānoa, Honolulu, HI 96822
  - code: CU
    address: Ecology and Evolutionary Biology, Universtity of Colorado, Boulder, CO 80309
  - code: NIAB
    address: Department of Crop Science and Production Systems, NIAB, Cambridge, CB3 0LE, UK
  - code: UCD
    address: Department of Plant Sciences, University of California, Davis, CA 95616
  - code: UWM
    address: Department of Botany, University of Wisconsin, Madison, WI 53706
# footnote:
#   - code: 1
#     text: "Current email address: \\href{mailto:cdmuir@wisc.edu}{cdmuir@wisc.edu}"
abstract:
  - "We will write abstract after Discussion and Conclusions are complete"
keywords:
  - amphistomy
  - Arabidopsis thaliana
  - CO$_2$ diffusion
  - finite element method
  - optimality
  - photosynthesis
  - stomata
fontsize: 12pt
competing_interests: |
  The authors declare no competing interests.
author_contributions: |
  JLW and CDM conceived of the project, analyzed data, and wrote the manuscript. GW provided data. TNB contributed to model development and helped edit the manuscript.
acknowledgements: |
  This work was supported by NSF Award 1929167 to CDM and NSF Award 2307341 to TNB.
bibliography: stomata-spacing.bib
link-citations: yes
output: 
  rticles::oup_article:
    latex_engine: xelatex
    oup_version: 1 # 1 = 2020 CTAN OUP CLS package; 0 = 2009 OUP CLS package
    # journal: "AoB PLANTS"
    document_style: "modern" # Can be contemporary, modern, traditional
    papersize: "large" # Can be large, medium, small
    #citation_package: "default" # Uncomment when using a CSL; default "natbib"
    namedate: TRUE # Set FALSE to use numeric refs; Default FALSE
    number_sections: FALSE # Uncomment to not number sections; default TRUE
    #number_lines: TRUE # Use `lineno` package - Default FALSE
    #number_lines_options: ["mathlines","switch"]  # Options for latex lineno package.
    onecolumn: TRUE # Uncomment for one column format; default FALSE
    extra_dependencies:
      - booktabs # to use with knitr::kable() example below
header-includes:
 - \usepackage[nomarkers,tablesfirst]{endfloat} # For figures and tables at end
 - \usepackage{lineno} # For line numbering
 - \linenumbers # For line numbering
 - \usepackage{hyperref} # For autoref
 - \renewcommand{\figureautorefname}{Fig.}
 - \usepackage[detect-none]{siunitx} # for \numrange{}{}
 - \sisetup{range-phrase = \text{--}}
 - \usepackage{caption} # for \captionsetup{}
 - \usepackage{newunicodechar,graphicx}
 - \DeclareRobustCommand{\okina}{\raisebox{\dimexpr\fontcharht\font`A-\height}{\scalebox{0.8}{`}}}
 - \newunicodechar{ʻ}{\okina}

@@

# Mean stomatal density ratio
mean_sr = single_surface_results |>
  mutate(rep = str_extract(image_number, "[0-9]+")) |>
  select(treatment, light, leaf_number, rep, surface, n_stomata) |>
  pivot_wider(values_from = "n_stomata", names_from = surface) |>
  mutate(sr = adaxial / (abaxial + adaxial)) |>
  pull(sr) |>
  mean()

@@

Plant material, growth conditions, and three-dimensional confocal imaging are described in @dow_disruption_2017. Briefly, Columbia (Col-0) ecotype of *Arabidopsis thaliana* (L.) Heynh. plants were grown in three different light environments: low light (PAR = 50 $\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$), medium light (100 $\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$), and high light (200 $\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$). PAR stands for photosynthetically active radiation. Seeds were surface-sterilized and stratified at 4°C for 3–5 d in 0.15% agarose solution and then sown directly into Pro-Mix HP soil (Premier Horticulture; Quakerstown, PA, USA) and supplemented with Scott's Osmocote Classic 14-14-14 fertilizer (Scotts-Sierra, Marysville, OH, USA). At 10–14 d, seedlings were thinned so only one seedling per container remained. Plants were grown to maturity in growth chambers where the conditions were as follows: 16 : 8 h, 22 : 20°C, day : night cycle. Imaging of the epidermis and internal leaf structures was performed using a Leica SP5 confocal microscope with the protocol developed by @wuyts_high-contrast_2010 with additional modification described in @dow_disruption_2017. We captured 132 images in total, making 66 abaxial-adaxial image pairs. We measured stomatal position and size using ImageJ [@schneider_nih_2012].

@@

Stomatal density of *Arabidopsis thaliana* varies among light treatments (ANOVA, $`r report_fstat(density_anova, 'light')`$) because the density is much greater in the high light treatment (\autoref{fig:density}). Density is consistently greater on abaxial leaf surfaces across all light treatments (ANOVA, $`r report_fstat(density_anova, 'surface')`$; \autoref{fig:density}). There is no evidence for an interaction between light treatment and surface (ANOVA, $`r report_fstat(density_anova, 'light:surface')`$). Leaves are amphistomatous with a mean stomatal density ratio of `r prettyNum(mean_sr, digits = 2)`.

### 2023-10-19 ms/ms.tex
fixed scale following Graham's email on 2023-10-19

% Create labels for Addresses if the are given in Elsevier format
   \MyNewLabel{ADDRTXT}{UHM}{%
  School of Life Sciences, University of Hawaiʻi at Mānoa, Honolulu, HI
96822%
 }
   \MyNewLabel{ADDRTXT}{CU}{%
  Ecology and Evolutionary Biology, Universtity of Colorado, Boulder, CO
80309%
 }
   \MyNewLabel{ADDRTXT}{NIAB}{%
  National Institute of Agricultural Botany, Cambridge, United Kingdom
CB3 0LE%
 }
   \MyNewLabel{ADDRTXT}{UCD}{%
  Department of Plant Sciences, University of California, Davis, CA
95616%
 }
   \MyNewLabel{ADDRTXT}{UWM}{%
  Department of Botany, University of Wisconsin, Madison, WI 53706%
 }

@@

\author[%
\addrlblbycode{NIAB}%
%
%
%
]{Graham Dow}

@@

{[}Graham - can you fill in details on growth conditions, image
aquisitiion, and image dimensions?{]}

\emph{Arabidopsis thaliana} plants were grown in three different light
environments: low light (PAR = 50
\(\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\)), medium light (100
\(\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\)), and high light (200
\(\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\)). PAR stands for
photosynthetically active radiation. Once leaves were mature, we
captured images of the abaxial and adaxial leaf surfaces using XXX. We
captured 132 images in total, making 66 abaxial-adaxial image pairs. We
measured stomatal position and size using ImageJ
\citep{schneider_nih_2012}.

@@

Stomatal density of \emph{Arabidopsis thaliana} the 132 leaf surfaces
measured range from 12 to 93 (units) with high light leaves ranging from
93 to 55 (units), medium light from 15 to 35 (units), and low light from
12 to 42 (units). Stomatal density varies among light treatments (ANOVA,
\(F_{2,126} = 681, P = 2.88 \times 10^{-68}\)) because the density is
much greater in the high light treatment (\autoref{fig:density}).
Density is consistently greater on abaxial leaf surfaces across all
light treatments (ANOVA, \(F_{1,126} = 44.2, P = 8.21 \times 10^{-10}\);
\autoref{fig:density}). There is no evidence for an interaction between
light treatment and surface (ANOVA,
\(F_{2,126} = 2.75 \times 10^{-2}, P = 0.973\)). Leaves are
amphistomatous with a mean stomatal density ratio of 0.45.

>>>>

% Create labels for Addresses if the are given in Elsevier format
   \MyNewLabel{ADDRTXT}{UHM}{%
  School of Life Sciences, University of Hawaiʻi at Mānoa, Honolulu, HI
96822%
 }
   \MyNewLabel{ADDRTXT}{CU}{%
  Ecology and Evolutionary Biology, Universtity of Colorado, Boulder, CO
80309%
 }
   \MyNewLabel{ADDRTXT}{NIAB}{%
  Department of Crop Science and Production Systems, NIAB, Cambridge,
CB3 0LE, UK%
 }
   \MyNewLabel{ADDRTXT}{UCD}{%
  Department of Plant Sciences, University of California, Davis, CA
95616%
 }
   \MyNewLabel{ADDRTXT}{UWM}{%
  Department of Botany, University of Wisconsin, Madison, WI 53706%
 }

@@

\author[%
\addrlblbycode{NIAB}%
%
%
%
]{Graham J. Dow}

@@

Plant material, growth conditions, and three-dimensional confocal
imaging are described in \citet{dow_disruption_2017}. Briefly, Columbia
(Col-0) ecotype of \emph{Arabidopsis thaliana} (L.) Heynh. plants were
grown in three different light environments: low light (PAR = 50
\(\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\)), medium light (100
\(\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\)), and high light (200
\(\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\)). PAR stands for
photosynthetically active radiation. Seeds were surface-sterilized and
stratified at 4°C for 3--5 d in 0.15\% agarose solution and then sown
directly into Pro-Mix HP soil (Premier Horticulture; Quakerstown, PA,
USA) and supplemented with Scott's Osmocote Classic 14-14-14 fertilizer
(Scotts-Sierra, Marysville, OH, USA). At 10--14 d, seedlings were
thinned so only one seedling per container remained. Plants were grown
to maturity in growth chambers where the conditions were as follows: 16
: 8 h, 22 : 20°C, day : night cycle. Imaging of the epidermis and
internal leaf structures was performed using a Leica SP5 confocal
microscope with the protocol developed by
\citet{wuyts_high-contrast_2010} with additional modification described
in \citet{dow_disruption_2017}. We captured 132 images in total, making
66 abaxial-adaxial image pairs. We measured stomatal position and size
using ImageJ \citep{schneider_nih_2012}.

@@

Stomatal density of \emph{Arabidopsis thaliana} varies among light
treatments (ANOVA, \(F_{2,126} = 681, P = 2.88 \times 10^{-68}\))
because the density is much greater in the high light treatment
(\autoref{fig:density}). Density is consistently greater on abaxial leaf
surfaces across all light treatments (ANOVA,
\(F_{1,126} = 44.2, P = 8.21 \times 10^{-10}\); \autoref{fig:density}).
There is no evidence for an interaction between light treatment and
surface (ANOVA, \(F_{2,126} = 2.75 \times 10^{-2}, P = 0.973\)). Leaves
are amphistomatous with a mean stomatal density ratio of 0.44.
